### 2014-02-01 Studies/Chicago_2012/Report/Chicago_Fire.tex
Update to Chicago writeup

On November 2, 2012, a fire in a 2 1/2 story wood frame residential structure claimed the life of a Chicago firefighter.  Changes in the ventilation flow path cause a rapid change in the fire dynamics resulting in fire and smoke penetration to an interior second floor hallway. NIST assisted with examining the fire dynamics of this incident. NIST performed computer simulations of the fire using the FDS and Smokeview, a visualization tool, to provide insight on the fire development and thermal conditions that may have existed in the residence during the fire. The specific objectives of the simulations detailed in this report are: 
\begin{enumerate}
\item To examine with physics-based calculations the effect of flow path on the thermal and fire conditions in this multi-story residential structure.
\item To provide visualization of fire behavior representative of what Chicago Fire Department members experienced during the course of their interior operations.
\end{enumerate}
This document describes the input and the results of the FDS (version 6.0.0) simulations.

@@

\chapter{Fire Dynamics Simulator and Smokeview}
FDS is a computational fluid dynamics (CFD) model that solves a form of the Navier-Stokes equations appropriate for low-speed, thermally driven flow with an emphasis on smoke and heat transport from fires.  Within a CFD model, the room or building is divided into small three-dimensional rectangular control volumes or computational cells.  The cells are contained together within one larger volume known as a computational domain.  The CFD model computes the density, velocity, temperature, pressure and species concentration of the gas in each cell.  Based on the laws of conservation of mass, momentum,  and energy, the model tracks the generation and movement of fire gasses.  FDS can utilize material properties of the furnishings, walls, floors, and ceilings to compute fire growth and spread.  One of the most important advantages of FDS is that it is verified and validated against fire test data to ensure that it provides the expected results, given sufficient input data \cite{FDS_Verification_Guide,FDS_Validation_Guide}.  A complete description of the FDS model is given in reference \cite{FDS_Tech_Guide}.

@@

\chapter{Model Input Parameters}
Input data from a number of sources must be collected and documented in order to simulate a fire with FDS or any other model. 
From the fire scene the following information must be gathered:
\begin{itemize}
\item The geometry of the building and the compartments being modeled.
\item The size and location of exterior and interior ventilation openings. 
\item Documentation and identification of the fuels used in the construction and finish of the building and their installed location
\item Documentation and identification of the fuels used as the contents of the building. 
\item Area of origin and approximate size of ignition source.
\item Documentation of fire damage to the building, fuel loads, victims or equipment.
\end{itemize}
From witnesses, first responders, reports, and recorded media such as fire ground radios or videos:
\begin{itemize}
\item Information on the timing of the fire development
\item Order and timing of ventilation opening to the outside
\item Weather conditions at the time of the fire
\end{itemize}
To develop the simulation, a timeline of events based on the collected information must be developed.   From this time line actions may be input into the model, such as venting a window at a given time.  Or the time line may be used in validating a model result.

@@

\section{Geometry}
\label{geom}
The structure is a 2 $\tfrac{1}{2}$ story residential structure with balloon frame construction and a wood gable roofing system. The structure has a 17.15 m (56 ft) by 6.15 m (20 ft) footprint. This results in a floor area of approximately 105 m$^2$ (1100 ft$^2$) per floor and a total living area of approximately 110 m$^2$ (2200 ft$^2$). Additionally there is a half-story attic that was used to house the structure's hot water heater. Interior walls were constructed from wood frame dimensioned lumber with 1.27 cm (0.5 in) gypsum board over top. There were two stairwells in the structure: the front stairwell connected the first and second levels and the rear stairwell (part of the enclosed porch) connected the first and second levels and provided access to the attic. Figs.~(\ref{fig:first_floor} and \ref{fig:second_floor}) show the interior dimensions of the first and second floor, respectively. The dimensions were found from on-scene measurements using a tape measure. Measurements were made to the nearest 1.27 cm (0.5 in). Based on difficulty of making some of the measurements due to the condition of the post-fire structure, the uncertainty in the measurements is estimated to be on the order of $\pm$ 1 \%. This would result in measurements that could change by up $\pm$ 17 cm ($\approx$ 7 in). When considering the uncertainties in the fire ignition time, ventilation, and peak heat release rate among other factors, the uncertainty in the overall structure geometry is not expected to be a major factor.

The model geometry was constructed based on the measurements taken at the structure. The resolution of the model is to the nearest centimeter with that the recognition that the resolution is ultimately limited by the mesh resolution which will be discussed in Section \ref{mesh}. Fig.~(\ref{fig:smv_interior}) shows Smokeview renderings of cutaways of the beta and delta sides of the structure interior.

\begin{figure}[h!]
\begin{tabular*}{\textwidth}{l@{\extracolsep{\fill}}r}
\includegraphics[width=.80\textwidth]{../Figures/interior_beta} \\
\includegraphics[width=.80\textwidth]{../Figures/interior_delta} \\
\end{tabular*}
\caption{3D Smokeview rendering of structure interior looking from the charlie side: to the beta side (top) and to the delta side (bottom)}
\label{fig:smv_interior}
\end{figure}

@@

Next, is determining the appropriate burnable area (exposed wood area) in the attic and in the second floor porch enclosure. Based on the floor plan of the structure, the underside of the roof is approximately 134 m$^2$ (1475 ft$^2$) plus there is 105 m$^2$ of floor area in the attic. Examination of post incident images showed the roof sustained more damage than the floor except for the floor area immediately over the back porch. That area will be considered as part of the porch fire. To estimate the attic fire size, we consider the area of the underside of the roof (134 m$^2$) and use an average heat release rate of wood of 65 kW/m$^2$. This results in a fire size of about 9 MW. To estimate the porch fire size, we consider the the four walls of porch enclosure (42 m$^2$) and the ceiling above the porch (13 m$^2$). The total area is 55 m$^2$ (592 ft$^2$). Using the same average heat release of 65 kW/m$^2$, the fire size on the porch is approximately 3.3 MW. The total fire size for this structure is estimated to be 12.3 MW. Since it is unknown when the fire started relative to notification of a fire, the attic fire starts at the beginning of the simulation and ramps up to 9 MW in under 10 seconds. The attic fire starts 40 seconds into the simulation and ramps to 3.3 MW in under 10 seconds.

\section{Vents}
\label{Vents}
This simulation includes vents that account for the approximate leakage of the building envelope and changes in ventilation due to a combination of fire department operations (opening doors) and fire acting on the structure (fire breaching walls,breaking windows,etc). The times of the ventilation changes are provided in Table \ref{tab:vents}.

@@

For the simulation in this study, we can estimate a measure of how well the flow field is resolved by the non-dimensional expression $D^*/\dx$. Here $D^*$ is a characteristic fire diameter, $\dx$ is the nominal size of a mesh cell, and $\dQ$ is the total heat release rate of the fire:
\begin{equation}
D^* = \left(
     \frac{\dQ}{\rho_\infty \, c_p \, T_\infty \, \sqrt{g} }
     \right)^\frac{2}{5} 
\label{eq:mesh}
\end{equation}   
From the FDS User Guide \cite{FDS_Users_Guide}, ``the characteristic fire diameter is related to the characteristic fire size via the
relation $Q^* = (D^*/D)^{5/2}$.'' $D$ is the physical diameter of the fire. Following Eq.~(\ref{eq:mesh}) and using 10 cm grid cells, the characteristic fire diameter to cell size ratio based on the 3.3 MW porch fire is 15.5 and for the 9 MW fire is 23. Based on validation work performed for the U.S. Nuclear Regulatory Commission, $D^*/\dx$ values ranged between 4 and 16 \cite{NUREG_1824}. The grid resolution used in the model fall within typical, engineering $D^*/\dx$ values.

>>>>

On November 2, 2012, a fire in a 2 1/2 story wood frame residential structure claimed the life of a Chicago firefighter.  Changes in the ventilation flow path caused a rapid change in the fire dynamics resulting in fire and smoke penetration to an interior second floor hallway. NIST assisted with examining the fire dynamics of this incident. NIST performed computer simulations of the fire using Fire Dynamics Simulator and Smokeview, to provide insight on the fire development and thermal conditions that may have existed in the residence during the fire. The specific objectives of the simulations detailed in this report are: 
\begin{enumerate}
\item To examine with physics-based calculations the effect of flow path on the thermal and fire conditions in this multi-story residential structure.
\item To provide visualization of fire behavior representative of what Chicago Fire Department members experienced during the course of their interior operations.
\end{enumerate}
This document describes the input and the results of the FDS (version 6.0.0) simulations.

@@

\chapter{Fire Dynamics Simulator and Smokeview}
FDS is a computational fluid dynamics (CFD) model that solves a form of the Navier-Stokes equations appropriate for low-speed, thermally driven flow with an emphasis on smoke and heat transport from fires.  Within a CFD model, the room or building is divided into small three-dimensional rectangular control volumes or computational cells.  The cells are contained together within one larger volume known as a computational domain.  The CFD model computes the density, velocity, temperature, pressure and species concentration of the gas in each cell.  Based on the laws of conservation of mass, momentum,  and energy, the model tracks the generation and movement of fire gasses. One of the most important advantages of FDS is that it is  mathematically verified \cite{FDS_Verification_Guide} and validated against fire test data to ensure that it provides the expected results, given sufficient input data \cite{FDS_Validation_Guide}.  A complete description of the FDS model is given in reference \cite{FDS_Tech_Guide}.

@@

\chapter{Model Input Parameters}
Input data from a number of sources must be collected and documented in order to simulate a fire with FDS or any other model. 
From the fire scene the following information must be gathered:
\begin{itemize}
\item The geometry of the building and the compartments being modeled.
\item The size and location of exterior and interior ventilation openings. 
\item Documentation and identification of the fuels used in the construction and finish of the building and their installed location
\item Documentation and identification of the fuels used as the contents of the building. 
\item Area of origin and approximate size of ignition source.
\item Documentation of fire damage to the building, fuel loads, victims or equipment.
\end{itemize}
From witnesses, first responders, reports, and recorded media such as fire ground radios or videos:
\begin{itemize}
\item Information on the timing of the fire development
\item Order and timing of ventilation opening to the outside
\item Weather conditions at the time of the fire
\end{itemize}
To develop the simulation, a timeline of events based on the collected information must be developed.   From this time line actions may be input into the model, such as venting a window at a given time.

@@

\section{Geometry}
\label{geom}
The structure is a 2 $\tfrac{1}{2}$ story residential structure with balloon frame construction and a wood gable roofing system. The structure has a 17.15 m (56 ft) by 6.15 m (20 ft) footprint. This results in a floor area of approximately 105 m$^2$ (1100 ft$^2$) per floor and a total living area of approximately 110 m$^2$ (2200 ft$^2$). Additionally there is a half-story attic that was used to house the structure's hot water heater. Interior walls were constructed from wood frame dimensioned lumber with 1.27 cm (0.5 in) gypsum board over top. There were two stairwells in the structure: the front stairwell connected the first and second levels and the rear stairwell (part of the enclosed porch) connected the first and second levels and provided access to the attic. Figs.~(\ref{fig:first_floor} and \ref{fig:second_floor}) show the interior dimensions of the first and second floor, respectively. The dimensions were found from on-scene measurements using a tape measure. Measurements were made to the nearest 1.27 cm (0.5 in). Based on difficulty of making some of the measurements due to the condition of the post-fire structure, the uncertainty in the measurements is estimated to be on the order of $\pm$ 1 \%. This would result in measurements that could change by up $\pm$ 17 cm ($\approx$ 7 in). When considering the uncertainties in the initial fire development, ventilation (vent times and vent areas), and peak heat release rate among other factors, the uncertainty in the overall structure geometry is not expected to be a major factor.

The resolution of the model geometry input is to the nearest centimeter with that the recognition that the resolution is ultimately limited by the mesh resolution which will be discussed in Section \ref{mesh}. Fig.~(\ref{fig:smv_interior}) shows 3D Smokeview renderings of cutaways of the beta and delta sides of the structure interior.

\begin{figure}[h!]
\begin{tabular*}{\textwidth}{l@{\extracolsep{\fill}}r}
\includegraphics[width=.80\textwidth]{../Figures/interior_beta} \\
\includegraphics[width=.80\textwidth]{../Figures/interior_delta} \\
\end{tabular*}
\caption{3D Smokeview rendering of structure interior looking from the charlie side to both the beta side (top) and to the delta side (bottom)}
\label{fig:smv_interior}
\end{figure}

@@

Next, is determining the appropriate burnable area (exposed wood area) in the attic and in the second floor porch enclosure. Based on the floor plan of the structure, the underside of the roof is approximately 134 m$^2$ (1475 ft$^2$) plus there is 105 m$^2$ of floor area in the attic. Examination of post incident images showed the roof sustained more damage than the floor except for the floor area immediately over the back porch. That area will be considered as part of the porch fire. To estimate the attic fire size, we consider the area of the underside of the roof (134 m$^2$) and use an average heat release rate of wood of 65 kW/m$^2$. This results in a fire size of about 9 MW. To estimate the porch fire size, we consider the the four walls of porch enclosure (42 m$^2$) and the ceiling above the porch (13 m$^2$). The total area is 55 m$^2$ (592 ft$^2$). Using the same average heat release of 65 kW/m$^2$, the fire size on the porch is approximately 3.3 MW. The total fire size for this structure is estimated to be 12.3 MW. Since it is unknown when the fire started relative to notification of a fire, the attic fire starts at the beginning of the simulation and ramps up to 9 MW in 8 seconds. The attic fire starts 40 seconds into the simulation and ramps to 3.3 MW in 8 seconds.

\section{Vents}
\label{Vents}
This simulation includes vents that account for the approximate leakage of the building envelope and changes in ventilation due to a combination of fire department operations (opening doors) and fire acting on the structure (fire breaching walls, breaking windows, etc). The times of the ventilation changes are provided in Table \ref{tab:vents}.

@@

For the simulation in this study, we can estimate a measure of how well the flow field is resolved by the non-dimensional expression $D^*/\dx$. Here $D^*$ is a characteristic fire diameter, $\dx$ is the nominal size of a mesh cell, and $\dQ$ is the total heat release rate of the fire:
\begin{equation}
D^* = \left(
     \frac{\dQ}{\rho_\infty \, c_p \, T_\infty \, \sqrt{g} }
     \right)^\frac{2}{5} 
\label{eq:mesh}
\end{equation}   
From the FDS User Guide \cite{FDS_Users_Guide}, ``the characteristic fire diameter is related to the characteristic fire size via the
relation $Q^* = (D^*/D)^{5/2}$.'' Here, $D$ is the physical diameter of the fire as input to the model. Following Eq.~(\ref{eq:mesh}) and using 10 cm grid cells, the characteristic fire diameter to cell size ratio based on the 3.3 MW porch fire is 15.5 and for the 9 MW fire is 23. Based on validation work performed for the U.S. Nuclear Regulatory Commission, $D^*/\dx$ values ranged between 4 and 16 \cite{NUREG_1824}. The grid resolution used in the model falls within typical, engineering $D^*/\dx$ values. As a result, all of the input model geometry lengths, ventilation openings (holes, doors, and windows), and fire areas will snap to the nearest 10 cm.
